Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-20 paper/main.tex
more semi-dynamic linking and refs

\section{Introduction}
The annotation of genes in DNA is fundamental for many downstream tasks, which 
depend on high quality annotations. Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{}, PASA \citep{} or those offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and it seems 
actual learning of underlying patterns is limited despite the fundamental ability 
to do so being present \citep{}.

@@

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant
species from the JGI Phytosome 13 database \citep{goodstein2012phytozome}; exact 
genomes used are list here \ref{suptab:source_data}. % add a supplemental table with names and versions
Data was downloaded on 29/03/2019 and 15/10/2019 for plants and animals respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

For both feasibility and iterative cross-species generalization validattion, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we do want to train with as much data as possible, but 
not all data has a good enough quality to enable a powerful generalization. It is 
also very desirable to have a broad spread phylogenetically as well as in genome 
size and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize, if the genomic patterns inside the data are 
too different from each other. It was also important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

\subsection{Metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
and further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric takes the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S1 and S2 for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes. %?? hmmm, but this is not a weighted mean across classes is it? Do introns not weigh more heavily?

@@

The hyperparameters were optimized by a combination of manual and automatic 
optimization. Automatic optimization was carried out by using either the TPE 
algorithm \citep{bergstra2011algorithms}, random search or grid search depending on 
the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant 
hyperparameters can be found in Table S3 and S4 or in the Helixer source code 
repository.

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. 
We compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section 6 of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section 3 of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. The figures in Section 4 of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two models each, specifically the model with the best precision and the  % please check / epoch model / version something
other with the best recall. The model with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} gene caller, the popular gene caller, Augustus.
Typical usage requires retraining Augustus for each species, with the exception of a
few lucky cases where a model for a close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create a training set and
therefore could train and evaluate only models without UTRs.

@@

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors; but here we were particularly concerned because
the {\it de novo} gene caller used by the references was Augustus for many species (...)
% todo list and cite examples (above and below)
and a fundamentally similar HMM-based tool for many more (...), which would potentially bias 
reference based comparisons of Helixer's and Augustus' performance.

@@

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC \citep{}, before and after trimming
with Trimmomatic \citep{}, and then mapped to the genome with Hisat2 \citep{}. 
Samtools \citep{} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{} and custom downstream 
processing below. MultiQC \citep{} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, particularly to
exonic regions; that had low 3' bias, normal FastQC and Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{sec:longer}\end{supsec}
\begin{supsec}\label{sec:overlapping}\end{supsec}
\begin{supsec}\label{sec:training_data}\end{supsec}

>>>>

\section{Introduction}
The annotation of genes in DNA is fundamental for many downstream tasks, which 
depend on high quality annotations. Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{cantarel2008maker}, PASA \citep{haas2003improving} or those 
offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and it seems 
actual learning of underlying patterns is limited despite the fundamental ability 
to do so being present \citep{}.

@@

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant
species from the JGI Phytosome 13 database \citep{goodstein2012phytozome}; exact 
genomes used are list here \ref{suptab:source_data}. % add a supplemental table with names and versions
Data was downloaded on 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used seperately during training as well as during evaluation.

@@

For both feasibility and iterative cross-species generalization validattion, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
and a variety of genome sizes and average gene length as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize, if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

\subsection{Metrics}
We mainly use two metrics to judge model performance. The {\it Genic F1} is our 
primary metric that provides the most comprehensive picture of annotation quality 
in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does 
not take UTR predictions into account. This is done for comparability with AUGUSTUS 
as further explained in Section \ref{sec:augustus}. Both metrics work by summing up 
True Positives (TP), False Positives (FP) and False Negatives (FN) of each 
considered class before calculating precision and recall from these values and 
combining the two into a F1 score. Neither metric takes the values of the 
intergenic class into account, as this class is very abundant and appears to be by 
far the easiest to predict. (See Table \ref{tab:statistics_all} for the class 
distribution and Tables S\ref{suptab:generalization_animals} 
and S\ref{suptab:generalization_plants} for a more in-depth report our model performance 
including the intergenic class). In the case of the Genic F1, this means that 
effectively only the TP of the intergenic class are disregarded. The metrics 
essentially provide a weighted mean of the performances in the considered classes,
with weights proportional to class frequency.

@@

The hyperparameters were optimized by a combination of manual and automatic 
optimization. Automatic optimization was carried out by using either the TPE 
algorithm \citep{bergstra2011algorithms}, random search or grid search depending on 
the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant 
hyperparameters can be found in Table S\ref{suptab:params} and 
S\ref{suptab:classweights} or in the Helixer source code repository.

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. 
We compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{sec:dcnn} of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section \ref{sec:longer} of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. The figures in Section \ref{sec:overlapping} of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

A model ensemble with 8 components was used to generate the final predictions for 
each species. For this, we performed 4 separate animal and plant training runs and 
selected two checkpoints each, specifically the checkpoint from the epoch with the best precision and the  % please check / epoch model / version something
other from the epoch with the best recall. The model checkpoint with the best Genic F1 was always best in one of 
those. This was done to increase the diversity of the model ensemble. As with the 
overlapping, the fusion of the 8 individual predictions was done by averaging the 
softmax values of each base pair prediction.

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} gene caller, the popular gene caller, Augustus \citep{stanke2003gene}.
Typical usage requires retraining Augustus for each species, with the exception of a
few lucky cases where a model for a sufficiently close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create a training set and
therefore could train and evaluate only models without UTRs.

@@

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors; but here we were particularly concerned because
the {\it de novo} gene caller used by the references was Augustus for many species 
(e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; {\it Asparagus officinalis}, 
\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and a different HMM-based tool for many more (e.g. {\it Citrus clementina}, \cite{wu2018genomics}; 
{\it Capsella grandiflora}, \cite{slotte2013capsella};,...), which would potentially bias 
reference based comparisons of Helixer's and Augustus' performance. % examples chosen alphabetically and skipping ons where I wasn't sure from the methods

@@

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC \citep{}, before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{Picard2019toolkit} and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, particularly to
exonic regions; that had low 3' bias, normal FastQC and Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{sec:generalization}\end{supsec}
\begin{supsec}\label{sec:hyperparameters}\end{supsec}
\begin{supsec}\label{sec:longer}\end{supsec}
\begin{supsec}\label{sec:overlapping}\end{supsec}
\begin{supsec}\label{sec:training_data}\end{supsec}
\begin{supsec}\label{sec:dcnn}\end{supsec}
\begin{supsec}\label{sec:rnaseq_eval}\end{supsec}
